Comparing Parallel Sorting Algorithms on a High Performance Computing Cluster Using OpenMP

### 2020-11-16 paper/Paper.tex
Updated paper

\section{Introduction}
Sorting is a fundamental problem to be solved in many algorithms and applications.
Any algorithm or application, for instance, that depends on having some ordering to its data will likely deal with sorting in some form. Because of its pervasiveness, solving the sorting problem efficiently is highly desirable. For this reason, many efficient sorting algorithms have been developed. Three of the most popular are Mergesort, Quicksort, and Heapsort. All three of these algorithms have time complexities of $O(nlog(n))$ in the average case. Mergesort and Heapsort also achieve this time bound in the worst case, while Quicksort can in rare situations exhibit a runtime of $O(n^2)$ in the worst case. Since the asymptotic lower bound for comparison-based sorting algorithms such as these is $\Omega(nlog(n))$, significantly improving the performance of them isn't feasible with serial computation. For this reason, we must turn to parallel computation to achieve significant performance gains. In order to take advantage of the performance increases enabled by parallel computing, parallelized versions of various sorting algorithms have been developed and studied. While theoretical analysis of parallel sorting algorithms is useful for understanding the asymptotic differences in the runtimes, determining and analysing the empirical performance of the algorithms on a specific computer architecture is valuable for making an informed choice about which algorithm to choose for that particular architecture and under various conditions (e.g., the number of CPUs available and the size of the input array to sort).
This study focuses on an empirical evaluation of parallelized versions of two of the most popular sorting algorithms: Quicksort and Mergesort. Heapsort is also analyzed as it is comparable to Quicksort and Mergesort, although it is not parallelized due to difficulties that we will discuss. The algorithms are parallelized with OpenMP so as to take advantage of multithreaded, shared-memory parallelism on Case Western Reserve's HPC Markov cluster.
The layout of the paper is as follows: In Section II, we give a brief overview of parallel computing and sorting algorithms. Section III...

@@

As mentioned before, all comparison based sorts have an asymptotic lower bound of $\Omega(nlog(n)$. 
Indeed,  Mergesort, and Heapsort achieve a $O(nlog(n))$ upper bound as well, and Quicksort behaves as $O(nlog(n))$ in the average case. 
Insertion sort, as a less advanced algorithm, has a upper bound of $O(n^2)$, as does Quicksort. \cite{cormen_introduction_2009} 
However, Quicksort's upper bound is rarely the case, and tends to have smaller constant factors then either Mergesort or Heapsort. \cite{hoare_algorithm_1961} % TODO wrong cite
Additionally, each algorithm is constructed differently, and some are more amendable to parallelization than others.

@@

\subsection{Mergesort}
Mergesort is a divide-and-conquer sorting algorithm that recursively divides an array into subarrays and merges them together in such a way that the original array is sorted. 
The operation of this procedure on an array of eight elements is depicted in Figure \ref{mrg}.  
\begin{figure}[h]
	\includegraphics[width=6cm]{merge.png} 
	\caption{Mergesort diagram from \cite{cormen_introduction_2009}}
	\label{mrg}
\end{figure}
Mergesort has a time complexity of $O(nlog(n))$ in both the average and worst case, although constant factors can make it worse than Quicksort in practice. 
Given the divide-and-conquer paradigm that Mergesort uses, it is a natural candidate for parallelization. Mergesort follows this paradigm as follows: First the input array (of size $n$) is recursively divided in two until there are $n$ single-element subarrays (divide). Each single-element array is trivially sorted (conquer). From there, the sorted subarrays are merged together, resulting in a single array that is equivalent to the input array but in sorted order (combine). Given the mechanics of the algorithm, there are two primary ways to parallelize Mergesort: (1) Parallelize the divide stage (i.e., the two recursive calls to the Mergesort procedure) and (2) Parallelize the combine stage (i.e., the Merge procedure). It is possible to implement either (1), (2), or both. According to \cite{cormen_introduction_2009}, only parallelizing the recursive calls to Mergesort will result in diminishing returns on performance as the number of processors grows to more than a few dozen. In order to efficiently scale up to hundreds of processors, parallelizing the Merge procedure is necessary because it becomes the performance bottleneck.
Our implementation of parallelized Mergesort parallelizes the recursive calls to Mergesort while implementing a serial Merge procedure. There are a couple of reasons for this design decision. First, the HPC architecture we are using (the Case Western Reserve Markov cluster) has well under one hundred CPU cores per node. Thus, parallel Mergesort with only the divide stage parallelized theoretically scales well within this core-count range. Second, parallelizing the Merge procedure proved to be difficult with OpenMP due to a lack of low-level control over threads.

@@

This suggests that there is significant overhead with smaller array sizes, but this is overcome by efficeicny gains on larger array sizes. 
\section{Conclusion}
% TODO these are what we expect to see...
As expected, the three $O(n log(n))$ algorithms perform better then Insertion sort. 
Due to the more parallel nature of Quicksort and Mergesort, these algorithms benefit more from parallization then the more sequential Heapsort. 
This is because the divide and conquer strategy is inherently more parallel, which should be taken into account when developing new algorithms to be run on parallel and distributed computing platforms. 
% TODO further findings from implementation
\bibliography{ref.bib}
\bibliographystyle{ieeetran}
\appendix
% TODO slap the code in here
\end{document}

>>>>

\section{Introduction}
Sorting is a fundamental problem to be solved in many algorithms and applications.
Any algorithm or application, for instance, that depends on having some ordering to its data will likely deal with sorting in some form. Because of its pervasiveness, solving the sorting problem efficiently is highly desirable. For this reason, many efficient sorting algorithms have been developed. Three of the most popular are Mergesort, Quicksort, and Heapsort. All three of these algorithms have time complexities of $O(nlog(n))$ in the average case. Mergesort and Heapsort also achieve this time bound in the worst case, while Quicksort can in rare situations exhibit a runtime of $O(n^2)$ in the worst case. Since the asymptotic lower bound for comparison-based sorting algorithms such as these is $\Omega(nlog(n))$, significantly improving the performance of them isn't feasible with serial computation. For this reason, we must turn to parallel computation to achieve significant performance gains. In order to take advantage of the performance increases enabled by parallel computing, parallelized versions of various sorting algorithms have been developed and studied. While theoretical analysis of parallel sorting algorithms is useful for understanding the asymptotic differences in the runtimes, determining and analysing the empirical performance of the algorithms on a specific computer architecture is valuable for making an informed choice about which algorithm to choose for that particular architecture and under various conditions (e.g., the number of CPUs available and the size of the input array to sort).
This study focuses on an empirical evaluation of parallelized versions of two of the most popular sorting algorithms: Mergesort and Quicksort. Heapsort is also analyzed as it is comparable to Mergesort and Quicksort, although it is not parallelized due to difficulties that we will discuss. The algorithms are parallelized with OpenMP so as to take advantage of multithreaded, shared-memory parallelism on Case Western Reserve's HPC Markov cluster.
The layout of the paper is as follows: In Section II, we give a brief overview of parallel computing and sorting algorithms. Section III...

@@

As mentioned before, all comparison based sorts have an asymptotic lower bound of $\Omega(nlog(n))$. 
Indeed,  Mergesort, and Heapsort achieve a $O(nlog(n))$ upper bound as well, and Quicksort behaves as $O(nlog(n))$ in the average case. 
Insertion sort, as a less advanced algorithm, has a upper bound of $O(n^2)$, as does Quicksort. \cite{cormen_introduction_2009} 
However, Quicksort's upper bound is rarely the case, and tends to have smaller constant factors then either Mergesort or Heapsort. \cite{hoare_algorithm_1961} % TODO wrong cite
Additionally, each algorithm is constructed differently, and some are more amendable to parallelization than others.

@@

\subsection{Mergesort}
Mergesort is a divide-and-conquer sorting algorithm that recursively divides an array into subarrays and merges them together in such a way that the original array is sorted. 
The operation of this procedure on an array of eight elements is depicted in Figure \ref{mrg}.  
\begin{figure}[h]
	\includegraphics[width=6cm]{merge.png} 
	\caption{Mergesort diagram from \cite{cormen_introduction_2009}}
	\label{mrg}
\end{figure}
Mergesort has a time complexity of $O(nlog(n))$ in both the average and worst case, although constant factors can make it worse than Quicksort in practice. 
Given the divide-and-conquer paradigm that Mergesort uses, it is a natural candidate for parallelization. Mergesort follows this paradigm as follows: First the input array (of size $n$) is recursively divided in two until there are $n$ single-element subarrays (divide). Each single-element array is trivially sorted (conquer). From there, the sorted subarrays are merged together, resulting in a single array that is equivalent to the input array but in sorted order (combine). Given the mechanics of the algorithm, there are two primary ways to parallelize Mergesort: (1) Parallelize the divide stage (i.e., the two recursive calls to the Mergesort procedure) and (2) Parallelize the combine stage (i.e., the Merge procedure). It is possible to implement either (1), (2), or both. According to \cite{cormen_introduction_2009}, only parallelizing the recursive calls to Mergesort will result in diminishing returns on performance as the number of processors grows to more than a few dozen. In order to efficiently scale up to hundreds of processors, parallelizing the Merge procedure is necessary because it becomes the performance bottleneck.
Our implementation of parallelized Mergesort parallelizes the recursive calls to Mergesort while implementing a serial Merge procedure. There are a couple of reasons for this design decision. First, the HPC architecture we are using (the Case Western Reserve Markov cluster) has well under one hundred CPU cores per node. Thus, parallel Mergesort with only the divide stage parallelized theoretically scales well within this core-count range. Second, parallelizing the Merge procedure proved to be difficult with OpenMP due to a lack of low-level control over threads. Our implementation of parallelized Mergesort is significantly based on Mergesort algorithms in Algorithms, Copyright (C) 2011  Atanas Radenski.

@@

This suggests that there is significant overhead with smaller array sizes, but this is overcome by efficeicny gains on larger array sizes. 
\section{Conclusion}
% TODO these are what we expect to see...
As expected, the three $O(n log(n))$ algorithms perform better then Insertion sort. 
Due to the more parallel nature of Quicksort and Mergesort, these algorithms benefit more from parallization then the more sequential Heapsort. 
This is because the divide and conquer strategy is inherently more parallel, which should be taken into account when developing new algorithms to be run on parallel and distributed computing platforms.

...

% TODO further findings from implementation
\bibliography{ref.bib}
\bibliographystyle{ieeetran}
\appendix
% TODO slap the code in here
\end{document}
